---
title: "Data Preparation: Denver Community Navigation Initiative (CNI)"
subtitle: "Synthetic Control Model — Lecture Dataset"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## 1. Background

The **Denver Community Navigation Initiative (CNI)** was a place-based policing reform
implemented in census tract **08031000800** (Five Points / Cole neighborhood). The policy
replaced traditional police responses with community health navigators for certain call types.

**Research question:** Did the CNI reduce crime and police stops in the treated neighborhood,
relative to a synthetic counterfactual?

This script produces two clean, quarterly panel datasets (2011 Q1 – 2021 Q4) for use in the
SCM lecture:

| Output file | Unit | Time | Key variables |
|---|---|---|---|
| `denver_crime_qtr.csv` | Census tract × quarter | 44 periods | `Property`, `Violence`, `Drug`, `Prostitution` (counts and per-1k) |
| `denver_stop_qtr.csv` | Census tract × quarter | 44 periods | `allstop`, `no_action`, `citation`, `arrest` (by subject/vehicle) |

**Key identifiers**

- Treatment unit: `FIPS = 8031000800`
- Treatment start: 2016 Q3 (`qtr = 2016.5` in `yearqtr` numeric encoding)
- Adjacent tracts (excluded from donor pool): identified programmatically via spatial
  adjacency (`st_touches`) — see Section 3

---

## 2. Setup

```{r libraries}
library(tidyverse)   # data wrangling
library(zoo)         # yearqtr / yearmon classes
library(sf)          # spatial operations
library(tigris)      # census tract shapefiles
library(tidycensus)  # ACS demographic data
```

```{r paths}
dir_raw   <- here::here("data_raw")
dir_clean <- here::here("data_clean")
dir.create(dir_clean, showWarnings = FALSE)

# ACS variables:  total pop, median HH income, race/ethnicity breakdown
acs_vars <- c("B19013_001", "B03002_001", "B03002_003",
              "B03002_004", "B03002_006", "B03002_012")

# CNI treated tract
FIPS_CNI <- 8031000800
# Adjacent tracts are computed spatially in Section 3 (FIPS_ADJACENT)

YEARS_CRIME <- 2011:2021   # crime data years to use
YEARS_STOP  <- 2011:2021   # stop data years to use
```

```{r census_key, eval=FALSE}
# Run once to install your Census API key permanently.
# Replace the string with your own key (https://api.census.gov/data/key_signup.html)
census_api_key("5c619b6f1f7097a9b1453b8cfe2e96d46dd7d577", install = TRUE)
```

---

## 3. Geocoding: Assigning Census Tract FIPS Codes

The raw data contains latitude/longitude coordinates but no census tract identifier.
We assign tracts using a **spatial join**:

1. Download 2019 Denver census tract boundaries via `tigris`
2. Convert each incident's coordinates to an `sf` point geometry
3. Join points to tract polygons — each point inherits the FIPS code of the polygon it
   falls within

> **Why 2019 tracts?** Census 2010 tract definitions apply through 2019; 2020 definitions
> take effect from 2020 onward. Using 2019 boundaries provides the best coverage for our
> 2011–2021 study window with minimal boundary change.

```{r load_tracts, results='hide'}
options(tigris_use_cache = TRUE)

tracts_sf <- tracts(state = "CO", county = "Denver", year = 2019, progress_bar = FALSE) %>%
  select(GEOID, geometry) %>%
  mutate(FIPS = as.numeric(GEOID)) %>%
  st_transform(crs = 4326)
```

```{r adjacent_tracts}
# Identify all census tracts that share a boundary with the CNI tract.
# st_touches() returns TRUE for polygons that share an edge or vertex but do not overlap.
# These tracts are excluded from the donor pool to avoid spillover contamination.

cni_geom      <- tracts_sf %>% filter(FIPS == FIPS_CNI)
touch_idx     <- st_touches(tracts_sf, cni_geom, sparse = FALSE)[, 1]
FIPS_ADJACENT <- tracts_sf$FIPS[touch_idx]

cat("Tracts spatially adjacent to CNI tract", FIPS_CNI,
    "(excluded from donor pool):\n")
print(FIPS_ADJACENT)
```

```{r geocode_fn}
# Geocode a data frame that has geo_lat / geo_lon columns.
# Returns the data frame with an added FIPS column.
geocode_to_tract <- function(df) {
  pts <- df %>%
    filter(!is.na(geo_lat), !is.na(geo_lon)) %>%
    st_as_sf(coords = c("geo_lon", "geo_lat"), crs = 4326, remove = FALSE)

  st_join(pts, tracts_sf["FIPS"], join = st_within) %>%
    st_drop_geometry() %>%
    filter(!is.na(FIPS))
}

# Add any missing columns (filled with 0L) so downstream selects never fail.
ensure_cols <- function(df, cols, fill = 0L) {
  missing <- setdiff(cols, names(df))
  if (length(missing) > 0) df[missing] <- fill
  df
}

# Linear interpolation of NAs within the pre-treatment period only.
# Applied per tract (inside group_by(FIPS)); post-period values are left as-is.
#   x:   numeric vector (one outcome column for one tract)
#   qtr: yearqtr vector of the same length
interp_pre <- function(x, qtr) {
  pre <- qtr < as.yearqtr("2016 Q3")
  x[pre] <- zoo::na.approx(x[pre], na.action = na.pass, rule = 2)
  x
}
```

---

## 4. ACS Demographic Data

We merge **2016 ACS 5-year estimates** as baseline demographic predictors. Using a single
pre-treatment year keeps the panel join straightforward and is standard practice for SCM
covariate construction.

```{r acs_pull, cache=TRUE}
acs_tract <- get_acs(
  geography = "tract",
  state     = "CO",
  county    = "Denver",
  year      = 2016,
  variables = acs_vars,
  output    = "wide",
  geometry  = TRUE
) %>%
  mutate(area_sq_miles = as.numeric(st_area(geometry)) * 0.00000038610) %>%
  transmute(
    FIPS      = as.numeric(GEOID),
    pop       = B03002_001E,
    med_inc   = B19013_001E,
    pct_color = (B03002_001E - B03002_003E) / B03002_001E,
    pop_density = pop / area_sq_miles
  ) %>%
  filter(pop > 0) %>%
  st_drop_geometry()
```

---

## 5. Crime Data

### 5.1 Process one year

```{r crime_process_fn}
process_crime_year <- function(year) {
  infile <- file.path(dir_raw, paste0("crime_", year, "_reduced.csv"))

  raw <- read.csv(infile) %>%
    mutate(
      date  = as.Date(date, "%d%b%Y"),
      qtr   = as.yearqtr(date),
      year2 = format(date, "%Y")
    ) %>%
    filter(year2 == year)          # drop any mis-dated rows

  geocode_to_tract(raw)
}
```

### 5.2 Aggregate across all years

Offense categories: **Property**, **Violence**, **Drug**, **Prostitution**.
Counts are aggregated at the **FIPS × quarter** level, then normalized per 1,000 residents.

```{r crime_aggregate, cache=TRUE}
crime_raw <- map_dfr(YEARS_CRIME, process_crime_year)

crime_qtr <- crime_raw %>%
  count(FIPS, qtr, offense) %>%
  pivot_wider(names_from = offense, values_from = n, values_fill = 0) %>%
  # Ensure expected columns exist even if an offense type was never observed
  ensure_cols(c("Property", "Violence", "Drug", "Prostitution")) %>%
  # Drop any residual unclassified codes if present
  select(FIPS, qtr, Property, Violence, Drug, Prostitution) %>%
  arrange(FIPS, qtr) %>%
  # Merge ACS demographics
  inner_join(acs_tract, by = "FIPS") %>%
  # Per-1k rates
  mutate(
    Property.k     = Property     * 1000 / pop,
    Violence.k     = Violence     * 1000 / pop,
    Drug.k         = Drug         * 1000 / pop,
    Prostitution.k = Prostitution * 1000 / pop,
    Allcrime       = Property + Violence + Drug + Prostitution,
    Allcrime.k     = Allcrime * 1000 / pop
  ) %>%
  # CNI classification
  mutate(
    group = case_when(
      FIPS == FIPS_CNI            ~ 1L,   # treated tract
      FIPS %in% FIPS_ADJACENT     ~ 2L,   # spatially adjacent tracts (excluded)
      TRUE                        ~ 3L    # donor tracts
    )
  ) %>%
  # Exclude adjacent tracts from the donor pool to avoid spillover contamination
  arrange(desc(group), FIPS, qtr)
```

### 5.3 Balanced panel filter

SCM requires a **complete time series** for every unit. Keep only tracts observed in all
44 quarters (2011 Q1 – 2021 Q4).

```{r crime_balance}
crime_panel <- crime_qtr %>%
  filter(qtr < "2022 Q1") %>%
  drop_na() %>%
  group_by(FIPS) %>%
    filter(n() == 44) %>%
  ungroup() %>%
  mutate(qtr = as.numeric(qtr))   # numeric encoding: e.g., 2011.00, 2011.25, ...
```

```{r crime_summary}
cat("Crime panel: ", n_distinct(crime_panel$FIPS), "tracts ×",
    n_distinct(crime_panel$qtr), "quarters\n")
cat("  Treated unit rows:", sum(crime_panel$FIPS == FIPS_CNI), "\n")
cat("  Adjacent tracts:", n_distinct(crime_panel$FIPS[crime_panel$group==2]), "\n")
cat("  Donor pool tracts:", n_distinct(crime_panel$FIPS[crime_panel$group==3]), "\n")
```

---

## 6. Stop Data

### 6.1 Process one year

Police stop outcomes are coded as:

| `call_disposition2` | Meaning |
|---|---|
| `0` | No enforcement action |
| `1` | Citation issued |
| `2` | Arrest made |

Stop type (`problem`): `1` = subject stop, `2` = vehicle stop.

```{r stop_process_fn}
process_stop_year <- function(year) {
  infile <- file.path(dir_raw, paste0("stop_", year, "_all.csv"))

  raw <- read.csv(infile) %>%
    mutate(
      date  = as.Date(date, "%d%b%Y"),
      qtr   = as.yearqtr(date),
      year2 = format(date, "%Y")
    ) %>%
    filter(year2 == year)

  geocode_to_tract(raw)
}
```

### 6.2 Aggregate across all years

```{r stop_aggregate, cache=TRUE}
stop_raw <- map_dfr(YEARS_STOP, process_stop_year)

stop_qtr <- stop_raw %>%
  count(FIPS, qtr, call_disposition2, problem) %>%
  pivot_wider(
    names_from  = c(call_disposition2, problem),
    values_from = n,
    values_fill = 0
  ) %>%
  # Standardise column names: <disposition>_<type>
  rename(
    no_action_sub = `0_1`,
    no_action_veh = `0_2`,
    citation_sub  = `1_1`,
    citation_veh  = `1_2`,
    arrest_sub    = `2_1`,
    arrest_veh    = `2_2`
  ) %>%
  # Ensure all six columns exist even if a combination was never observed
  ensure_cols(c("no_action_sub", "no_action_veh",
                "citation_sub",  "citation_veh",
                "arrest_sub",    "arrest_veh")) %>%
  mutate(
    no_action = no_action_sub + no_action_veh,
    citation  = citation_sub  + citation_veh,
    arrest    = arrest_sub    + arrest_veh,
    allstop   = no_action + citation + arrest,
    allstop_sub = no_action_sub + citation_sub + arrest_sub,
    allstop_veh = no_action_veh + citation_veh + arrest_veh
  ) %>%
  arrange(FIPS, qtr) %>%
  inner_join(acs_tract, by = "FIPS") %>%
  mutate(
    allstop.k   = allstop   * 1000 / pop,
    no_action.k = no_action * 1000 / pop,
    citation.k  = citation  * 1000 / pop,
    arrest.k    = arrest    * 1000 / pop
  ) %>%
  mutate(
    group = case_when(
      FIPS == FIPS_CNI         ~ 1L,
      FIPS %in% FIPS_ADJACENT  ~ 2L,
      TRUE                     ~ 3L
    )
  ) %>%
  arrange(desc(group), FIPS, qtr)
```

### 6.3 Balanced panel filter

```{r stop_balance}
stop_panel <- stop_qtr %>%
  filter(qtr < "2022 Q1") %>%
  drop_na() %>%
  group_by(FIPS) %>%
    filter(n() == 44) %>%
  ungroup() %>%
  mutate(qtr = as.numeric(qtr))
```

```{r stop_summary}
cat("Stop panel: ", n_distinct(stop_panel$FIPS), "tracts ×",
    n_distinct(stop_panel$qtr), "quarters\n")
cat("  Treated unit rows:", sum(stop_panel$FIPS == FIPS_CNI), "\n")
cat("  Adjacent tracts:", n_distinct(stop_panel$FIPS[stop_panel$group ==2]), "\n")
cat("  Donor pool tracts:", n_distinct(stop_panel$FIPS[stop_panel$group ==3]), "\n")
```


## 7. Export Clean Datasets

Both datasets are saved in two formats:

- **CSV** — portable, human-readable, easy to share
- **RData** — preserves R object classes (e.g., `yearqtr` if needed), loads instantly with `load()`

```{r export}
# --- CSV ---
write_csv(crime_panel, file.path(dir_clean, "denver_crime_qtr.csv"))
write_csv(stop_panel,  file.path(dir_clean, "denver_stop_qtr.csv"))

# --- RData (both objects in a single file for convenience) ---
save(crime_panel, stop_panel,
     file = file.path(dir_clean, "denver_synth_data.RData"))

cat("Saved CSV:\n",
    " ", file.path(dir_clean, "denver_crime_qtr.csv"), "\n",
    " ", file.path(dir_clean, "denver_stop_qtr.csv"),  "\n\n",
    "Saved RData:\n",
    " ", file.path(dir_clean, "denver_synth_data.RData"), "\n\n",
    "To reload in a future session:\n",
    '  load("data_clean/denver_synth_data.RData")  # restores crime_panel and stop_panel\n')
```

---

## 9. Data Dictionary

### Output files in `data_clean/`

| File | Format | Contents |
|---|---|---|
| `denver_crime_qtr.csv` | CSV | Crime panel |
| `denver_stop_qtr.csv` | CSV | Stop panel |
| `denver_synth_data.RData` | RData | Both `crime_panel` and `stop_panel` objects |

### `denver_crime_qtr.csv`

| Column | Type | Description |
|---|---|---|
| `FIPS` | numeric | 10-digit census tract FIPS (leading zero dropped) |
| `qtr` | numeric | Quarter in `yearqtr` numeric format (e.g., 2011.00 = 2011 Q1, 2011.25 = 2011 Q2) |
| `Property` | integer | Quarterly property crime count |
| `Violence` | integer | Quarterly violent crime count |
| `Drug` | integer | Quarterly drug crime count |
| `Prostitution` | integer | Quarterly prostitution crime count |
| `Allcrime` | integer | Sum of all four crime types |
| `*.k` | numeric | Per-1,000 residents version of count columns |
| `pop` | numeric | Total tract population (ACS 2016) |
| `med_inc` | numeric | Median household income (ACS 2016) |
| `pct_color` | numeric | Share of residents who are non-white (ACS 2016) |
| `group` | integer | 1 = treated tract, 2 = adjacent tract, 3 = donor tract |

### `denver_stop_qtr.csv`

| Column | Type | Description |
|---|---|---|
| `FIPS` | numeric | Census tract FIPS |
| `qtr` | numeric | Quarter (same encoding as crime data) |
| `no_action_sub / _veh` | integer | Stops with no enforcement action, by subject / vehicle |
| `citation_sub / _veh` | integer | Stops resulting in citation, by subject / vehicle |
| `arrest_sub / _veh` | integer | Stops resulting in arrest, by subject / vehicle |
| `no_action / citation / arrest` | integer | Combined (subject + vehicle) totals |
| `allstop` | integer | Total quarterly stops |
| `allstop_sub / _veh` | integer | Total stops by type |
| `*.k` | numeric | Per-1,000 residents version of `allstop`, `no_action`, `citation`, `arrest` |
| `pop / med_inc / pct_color` | numeric | ACS 2016 demographics (same as crime data) |
| `group` | integer | 1 = treated tract, 2 = adjacent tract, 3 = donor tract |
