---
title: "Synthetic Control Model: Denver CNI Case Study"
subtitle: "Effects on Crime and Police Stops"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

---

## 1. Setup

```{r libraries}
library(tidyverse)   # data wrangling and ggplot
library(tidysynth)   # synthetic control model
library(knitr)       # tables
library(kableExtra)  # table formatting
```

```{r constants}
# Treated unit
FIPS_CNI <- 8031000800          # CNI census tract (leading zero dropped when numeric)

# Treatment timing: de facto CNI launch = 2016 Q3 (yearqtr numeric encoding = 2016.5)
I_TIME <- 2016.5

# Key dates to mark on every trend plot
KEY_DATES <- c(
  "De jure start\n(2014 Q1)"  = 2014.0,
  "De facto start\n(2016 Q3)" = 2016.5,
  "Project end\n(2018 Q3)"    = 2018.5
)

# Output folder
dir_out <- here::here("outputs")
dir.create(dir_out, showWarnings = FALSE)
```

---

## 2. Load and Prepare Data

The clean quarterly panel datasets were produced by `0100_dataprep.Rmd`.
Each file is a **balanced panel**: every census tract has exactly **44 quarters**
(2011 Q1 – 2021 Q4). `CNI == 1` identifies the treated tract; `CNI == 3` all
potential donor tracts.

```{r load_data}
crime_panel_raw <- read_csv(here::here("data_clean", "denver_crime_qtr.csv"),
                        show_col_types = FALSE) 
stop_panel_raw  <- read_csv(here::here("data_clean", "denver_stop_qtr.csv"),
                        show_col_types = FALSE) 

# Remove adjacent tracks
crime_panel <- crime_panel_raw %>%
  filter(group!=2) 

stop_panel <- stop_panel_raw %>%
  filter(group!=2) 

# Rename 'citation' to avoid conflict with the base R citation() function.
stop_panel <- stop_panel %>% 
  rename(n_citation = citation,
         n_citation_sub = citation_sub,
         n_citation_veh = citation_veh)

cat("Crime panel:", n_distinct(crime_panel$FIPS), "tracts ×",
    n_distinct(crime_panel$qtr), "quarters\n")
cat("Stop panel: ", n_distinct(stop_panel$FIPS),  "tracts ×",
    n_distinct(stop_panel$qtr),  "quarters\n")
```

---

## 3. Descriptive Analysis

Before running any model, we examine the raw trends.
**What to look for:** the CNI tract (coloured line) and the donor pool average
should move together **before** the treatment date — this is the pre-treatment
parallel trends assumption that gives the SCM its credibility.

### 3.1 Crime trends
```{r desc_crime, fig.width = 11, fig.height = 4}
p_crime_desc <- crime_panel %>%
  mutate(group = ifelse(FIPS == FIPS_CNI, "CNI Tract", "Donor Tracts (avg)")) %>%
  group_by(group, qtr) %>%
  summarise(Allcrime = mean(Allcrime, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = qtr, y = Allcrime, color = group)) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = KEY_DATES, linetype = "dashed",
             color = "black", linewidth = 0.4) +
  annotate("text", x = KEY_DATES + 0.1, y = Inf,
           label = names(KEY_DATES), hjust = 0, vjust = 1.4, size = 2.6) +
  scale_color_manual(values = c("CNI Tract" = "firebrick",
                                "Donor Tracts (avg)" = "grey50")) +
  scale_x_continuous(breaks = 2011:2021) +
  labs(title = "Quarterly crime counts: CNI tract vs. donor tract average",
       x = NULL, y = "Mean crime count", color = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")

p_crime_desc
ggsave(file.path(dir_out, "desc_crime_trends.png"),
       p_crime_desc, width = 11, height = 4, dpi = 300)
```

### 3.2 Police stop trends
```{r desc_stop, fig.width = 11, fig.height = 4}
p_stop_desc <- stop_panel %>%
  mutate(group = ifelse(FIPS == FIPS_CNI, "CNI Tract", "Donor Tracts (avg)")) %>%
  group_by(group, qtr) %>%
  summarise(allstop = mean(allstop, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = qtr, y = allstop, color = group)) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = KEY_DATES, linetype = "dashed",
             color = "black", linewidth = 0.4) +
  annotate("text", x = KEY_DATES + 0.1, y = Inf,
           label = names(KEY_DATES), hjust = 0, vjust = 1.4, size = 2.6) +
  scale_color_manual(values = c("CNI Tract" = "steelblue",
                                "Donor Tracts (avg)" = "grey50")) +
  scale_x_continuous(breaks = 2011:2021) +
  labs(title = "Quarterly police stops: CNI tract vs. donor tract average",
       x = NULL, y = "Mean stop count", color = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")

p_stop_desc
ggsave(file.path(dir_out, "desc_stop_trends.png"),
       p_stop_desc, width = 11, height = 4, dpi = 300)
```

---

## 4. The Synthetic Control Method — Step-by-Step

The **Synthetic Control Method (SCM)** builds a weighted average of untreated
("donor") units whose pre-treatment trajectory best matches the treated unit.
After treatment, the weighted average serves as a counterfactual: what would
have happened without the intervention? The gap between observed and synthetic
is the estimated treatment effect.

We walk through each step using **all police stops** as the running example.

---

### Step 1 — Declare the panel structure

`synthetic_control()` registers the panel dimensions: which column is the unit
ID, which is time, which unit is treated, and when treatment starts.

```{r synth_step1}
synth_step1 <- stop_panel %>%
  synthetic_control(
    outcome           = allstop,  # outcome variable
    unit              = FIPS,     # unit identifier (census tract)
    time              = qtr,      # time identifier (numeric yearqtr)
    i_unit            = FIPS_CNI, # treated unit
    i_time            = I_TIME,   # treatment start: 2016 Q3 = 2016.5
    generate_placebos = FALSE     # set TRUE later for inference
  )
```

---

### Step 2 — Specify pre-treatment predictors

Predictors tell the algorithm which characteristics to match between the treated
unit and the synthetic control. We use:

1. **Demographic characteristics** at 2015 (the last full pre-treatment year):
   median household income (nominal, in $1,000s), share of non-white residents, and population density.
   Income is expressed in thousands so its scale is comparable to the other predictors;
   the underlying data remain in nominal dollars.
2. **Pre-treatment outcome mean** over 2011–2015: the most important predictor —
   matching the level of the outcome before treatment.

```{r synth_step2}
synth_step2 <- synth_step1 %>%

  # Predictor 1: tract demographics at 2015
  generate_predictor(
    time_window = 2015,
    med_inc_k   = med_inc / 1000,  # median income in $1,000s (nominal)
    pct_color   = pct_color,       # share non-white
    pop         = pop,             # population population density
    pop_density = pop_density      # population population density
  ) %>%

  # Predictor 2: average outcome over the full pre-treatment period
  generate_predictor(
    time_window = seq(2011, 2015.75, 0.25),
    outcome_pre = mean(allstop, na.rm = TRUE)
  )
```

---

### Step 3 — Estimate donor weights

`generate_weights()` solves a constrained optimisation problem: find non-negative
weights $w_j \geq 0$ that sum to 1 and minimise the weighted distance between the
treated unit's predictors and the weighted average of donor predictors.

```{r synth_step3}
synth_step3 <- synth_step2 %>%
  generate_weights(
    optimization_window = seq(2011, 2015.75, 0.25)  # pre-treatment window
  )
```

---

### Step 4 — Construct the synthetic control

`generate_control()` applies the estimated weights to the **full** time series
(pre and post) to produce the counterfactual trajectory.

```{r synth_step4}
synth_allstop <- synth_step3 %>%
  generate_control()
```

---

### Step 5 — Inspect the fit

A valid synthetic control must track the treated unit closely **before** treatment.
We check this in three ways.

**A. Predictor balance** — do the predictors match?

```{r balance_table}
synth_allstop %>%
  grab_balance_table() %>%
  kable(digits = 2,
        caption = "Predictor balance: CNI tract vs. synthetic control (all stops)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

**B. Donor weights** — which tracts contribute to the synthetic control?

```{r unit_weights}
synth_allstop %>%
  grab_unit_weights() %>%
  filter(weight > 0.01) %>%
  arrange(desc(weight)) %>%
  kable(digits = 4, caption = "Donor weights > 1% (all stops)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

**C. Trend plot** — does the synthetic control track the pre-treatment path?

```{r walkthrough_plot, fig.width = 11, fig.height = 5}
synth_allstop %>%
  plot_trends(time_window = seq(2011, 2021.75, 0.25)) +
  geom_vline(xintercept = KEY_DATES, linetype = "dashed",
             color = "black", linewidth = 0.4) +
  scale_x_continuous(breaks = 2011:2021) +
  labs(
    title   = "All police stops: CNI tract (observed) vs. synthetic control",
    caption = "Dashed lines: de jure start (2014 Q1) | de facto start (2016 Q3) | project end (2018 Q3)",
    x = "Quarter", y = "Quarterly stops"
  ) +
  theme_minimal(base_size = 11)
```

---

## 5. Helper Functions

Now that we understand the steps, we encapsulate them into two reusable
functions — one for fitting, one for plotting. The same four steps from
Section 4 run inside `run_synth()` every time we call it.

```{r helpers}
# Fit the full SCM pipeline for one outcome variable.
# Returns a fitted tidysynth object ready for plotting and inference.
run_synth <- function(data, outcome_var) {
  data %>%
    synthetic_control(
      outcome           = {{ outcome_var }},
      unit              = FIPS,
      time              = qtr,
      i_unit            = FIPS_CNI,
      i_time            = I_TIME,
      generate_placebos = FALSE
    ) %>%
    generate_predictor(
      time_window = 2015,
      med_inc_k = med_inc / 1000,
      pct_color = pct_color,
      pop         = pop,             
      pop_density = pop_density
    ) %>%
    generate_predictor(
      time_window = seq(2011, 2015.75, 0.25),
      outcome_pre = mean({{ outcome_var }}, na.rm = TRUE)
    ) %>%
    generate_weights(optimization_window = seq(2011, 2015.75, 0.25)) %>%
    generate_control()
}

# Plot observed vs. synthetic trend and save to outputs/.
plot_synth_trends <- function(synth_obj, y_label, title, filename) {
  p <- synth_obj %>%
    plot_trends(time_window = seq(2011, 2021.75, 0.25)) +
    geom_vline(xintercept = KEY_DATES, linetype = "dashed",
               color = "black", linewidth = 0.4) +
    scale_x_continuous(breaks = 2011:2021) +
    labs(
      title   = title,
      caption = "Dashed lines: de jure start (2014 Q1) | de facto start (2016 Q3) | project end (2018 Q3)",
      x = "Quarter", y = y_label
    ) +
    theme_minimal(base_size = 11)
  print(p)
  ggsave(file.path(dir_out, filename), p, width = 11, height = 5, dpi = 300)
  invisible(p)
}
```

---

## 6. SCM Results — Police Stops

```{r stop_allstop, fig.width=11, fig.height=5}
synth_allstop <- run_synth(stop_panel, allstop)

plot_synth_trends(synth_allstop,
  y_label  = "Quarterly stops",
  title    = "SCM: All police stops — CNI tract vs. synthetic control",
  filename = "synth_stop_allstop.png")
```

```{r stop_noaction, fig.width=11, fig.height=5}
synth_no_action <- run_synth(stop_panel, no_action)

plot_synth_trends(synth_no_action,
  y_label  = "Stops (no enforcement action)",
  title    = "SCM: Stops with no enforcement action",
  filename = "synth_stop_no_action.png")
```

```{r stop_citation, fig.width=11, fig.height=5}
synth_citation <- run_synth(stop_panel, n_citation)

plot_synth_trends(synth_citation,
  y_label  = "Stops (citation issued)",
  title    = "SCM: Stops resulting in a citation",
  filename = "synth_stop_citation.png")
```

```{r stop_arrest, fig.width=11, fig.height=5}
synth_arrest <- run_synth(stop_panel, arrest)

plot_synth_trends(synth_arrest,
  y_label  = "Stops (arrest made)",
  title    = "SCM: Stops resulting in an arrest",
  filename = "synth_stop_arrest.png")
```

---

## 7. SCM Results — Crime

```{r crime_allcrime, fig.width=11, fig.height=5}
synth_allcrime <- run_synth(crime_panel, Allcrime)

plot_synth_trends(synth_allcrime,
  y_label  = "Quarterly crime incidents",
  title    = "SCM: All crimes — CNI tract vs. synthetic control",
  filename = "synth_crime_allcrime.png")
```

```{r crime_property, fig.width=11, fig.height=5}
synth_property <- run_synth(crime_panel, Property)

plot_synth_trends(synth_property,
  y_label  = "Property crimes",
  title    = "SCM: Property crimes",
  filename = "synth_crime_property.png")
```

```{r crime_violence, fig.width=11, fig.height=5}
synth_violence <- run_synth(crime_panel, Violence)

plot_synth_trends(synth_violence,
  y_label  = "Violent crimes",
  title    = "SCM: Violent crimes",
  filename = "synth_crime_violence.png")
```

```{r crime_drug, fig.width=11, fig.height=5}
synth_drug <- run_synth(crime_panel, Drug)

plot_synth_trends(synth_drug,
  y_label  = "Drug crimes",
  title    = "SCM: Drug crimes",
  filename = "synth_crime_drug.png")
```

---

## 8. Treatment Effect Summary

For each outcome we report the **average post-treatment gap**:
$$\text{Gap} = \bar{Y}^{\text{observed}} - \bar{Y}^{\text{synthetic}}$$
A negative gap means the CNI tract fell *below* its counterfactual — a reduction
attributable to the intervention.

```{r compute_gap}
# Extract the mean observed, mean synthetic, and gap for the post-treatment period.
compute_gap <- function(synth_obj, label) {
  synth_obj %>%
    grab_synthetic_control() %>%        # columns: time_unit, real_y, synth_y
    filter(time_unit >= I_TIME) %>%
    summarise(
      Outcome         = label,
      `Avg Observed`  = round(mean(real_y,           na.rm = TRUE), 1),
      `Avg Synthetic` = round(mean(synth_y,          na.rm = TRUE), 1),
      `Avg Gap`       = round(mean(real_y - synth_y, na.rm = TRUE), 1),
      `% Change`      = round(mean(real_y - synth_y, na.rm = TRUE) /
                              mean(synth_y,          na.rm = TRUE) * 100, 1)
    )
}
```

### 8.1 Crime

```{r effect_table_crime}
bind_rows(
  compute_gap(synth_allcrime, "All crimes"),
  compute_gap(synth_property, "Property crimes"),
  compute_gap(synth_violence, "Violent crimes"),
  compute_gap(synth_drug,     "Drug crimes")
) %>%
  kable(caption = "Treatment effects on crime (post-2016 Q3 quarterly average)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) %>%
  footnote(general = "Gap = Observed − Synthetic. Negative = reduction in CNI tract.")
```

### 8.2 Police Stops

```{r effect_table_stop}
bind_rows(
  compute_gap(synth_allstop,   "All stops"),
  compute_gap(synth_no_action, "No-action stops"),
  compute_gap(synth_citation,  "Citation stops"),
  compute_gap(synth_arrest,    "Arrest stops")
) %>%
  kable(caption = "Treatment effects on police stops (post-2016 Q3 quarterly average)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) %>%
  footnote(general = "Gap = Observed − Synthetic. Negative = reduction in CNI tract.")
```

---

## 9. Placebo (In-Space) Tests

### What is a placebo test?

We ask: *"What if we ran the same SCM for every donor tract, pretending each
was the treated unit?"*

Because donor tracts were **not** treated, we would expect small post-treatment
gaps for them. If the CNI tract's gap stands out as unusually large, it provides
evidence that the gap is real — not just noise.

**Inference:** the Fisher's exact p-value is the share of units (including
the treated unit) whose post/pre MSPE ratio is at least as large as the
treated unit's.

```{r placebo_helpers}
# Same as run_synth() but enables the placebo runs.
run_synth_placebo <- function(data, outcome_var) {
  data %>%
    synthetic_control(
      outcome           = {{ outcome_var }},
      unit              = FIPS,
      time              = qtr,
      i_unit            = FIPS_CNI,
      i_time            = I_TIME,
      generate_placebos = TRUE
    ) %>%
    generate_predictor(
      time_window = 2015,
      med_inc_k = med_inc / 1000,
      pct_color = pct_color,
      pop       = pop,
      pop_density = pop_density
    ) %>%
    generate_predictor(
      time_window = seq(2011, 2015.75, 0.25),
      outcome_pre = mean({{ outcome_var }}, na.rm = TRUE)
    ) %>%
    generate_weights(optimization_window = seq(2011, 2015.75, 0.25)) %>%
    generate_control()
}


# Plot placebo paths and save.
plot_placebo <- function(synth_obj, title, filename) {
  p <- synth_obj %>%
    plot_placebos(time_window = seq(2011, 2021.75, 0.25)) +
    geom_vline(xintercept = KEY_DATES, linetype = "dashed",
               color = "black", linewidth = 0.4) +
    scale_x_continuous(breaks = 2011:2021) +
    labs(
      title   = title,
      caption = "Grey lines = donor placebos. Coloured line = CNI tract.",
      x = "Quarter"
    ) +
    theme_minimal(base_size = 11)
  print(p)
  ggsave(file.path(dir_out, filename), p, width = 11, height = 5, dpi = 300)
  invisible(p)
}
```

### 9.1 Placebo: No-action

```{r placebo_allstop, fig.width=11, fig.height=5}
synth_no_action_placebo <- run_synth_placebo(stop_panel, no_action)

plot_placebo(synth_allstop_placebo,
  title    = "Placebo test: All police stops",
  filename = "placebo_stop_noaction.png")
```

```{r placebo_allstop_sig}
synth_no_action_placebo %>%
  grab_significance() %>%
  arrange(desc(mspe_ratio)) %>%
  kable(digits = 3,
        caption = "MSPE ratio ranking — No action (CNI tract vs. donor placebos)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

### 9.2 Placebo: Drug crime

```{r placebo_allcrime, fig.width=11, fig.height=5}
synth_drug_placebo <- run_synth_placebo(crime_panel, Drug)

plot_placebo(synth_drug_placebo,
  title    = "Placebo test: Drug crimes",
  filename = "placebo_crime_drug.png")
```

```{r placebo_allcrime_sig}
synth_drug_placebo %>%
  grab_significance() %>%
  arrange(desc(mspe_ratio)) %>%
  kable(digits = 3,
        caption = "MSPE ratio ranking — Drug crimes (CNI tract vs. donor placebos)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

---

## 10. Outputs

```{r summary, echo=FALSE}
cat("Files saved to:", dir_out, "\n\n")
cat(paste(" -", list.files(dir_out)), sep = "\n")
```
